Archaic introgression and the distribution of shared functional variation under stabilizing selection

### 2023-11-21 doc/manuscript.tex
finished drafting results

\subsection*{Expected genetic variance with linkage}

@@

The strength of the deviation of $V_G$ between models with and without linkage
depends on a number of factors. The total mutation rate affects not only the
polygenicity, with higher mutation rates causing higher number of segregating
alleles, but also the mean recombination rate between those alleles, as they
will be more or less densely distributed in the genome. The distribution of
effect sizes plays an important role, as seen by the opposite trends in $V_G$
relative to unlinked predictions, which are most apparent with high mutation
rates. There appear to be complex dynamics involving the Bulmer Effect
\citep{bulmer1971effect} and interference between selected alleles
\citep{hill1966effect}, which warrants careful investigation.

@@

\begin{itemize}
    \item Because introgressed trait-affecting alleles are selected against as the
        minor allele, linkage will also remove introgressed ancestry segments in the
        regions surrounding the selected alleles
    \item The expected reduction in introgressed ancestry proportions surrounding
        functional alleles will depend on the effect size of the trait-affecting allele
        and the probability of recombination surrounding that site
    \item We can consider first a deterministic model of the dynamics of introgressed
        allele frequencies at and around an introgressed functional allele -- this
        model considers a single trait-affecting locus, ignoring any effects of
        interference between selected alleles
        \begin{itemize}
            \item Consider a site with fixed differences between the two populations
                (the derived allele is fixed in one, lost in the other)
            \item At the functional site itself, with admixture proportion $f$ from the
                minor parental ancestry, the initial frequency of the derived allele
                is $p_0=f$ or $p_0 = 1-f$, if the derived allele was fixed in the 
                major parental ancestry population
            \item In one generation, the expected allele frequency at the selected locus
                is \[p_{t+1} \approx p_t - s p_t(1-p_t)(1-2p_t),\] where $s=a^2/2V_S$
            \item As discussed earlier, if $p_0<1/2$, $p_t\rightarrow0$, and if $p_0>1/2$,
                $p_t\rightarrow1$ as $t\rightarrow\infty$
            \item Consider a neutral locus separated from the selected locus by
                recombination probability $r$. Initially, the expected frequency
                of introgressed ancestry at that locus is $q_0=f$
            \item The linked ancestry frequency will change due to selection acting on
                the trait-affecting locus, depending on the strength of selection
                at that locus and the level of linkage disequilibrium between
                the two loci
            \item $D=Cov(p,q)$ is the standard covariance measure of LD between loci,
                and $q$ is expected to change as \[q_{t+1} \approx q_t - s D_t(1-2p_t)\]
            \item $D$ also changes over time, due to both selection and recombination,
                so that \[D_{t+1} \approx D_t - r D_t - s D_t (1-2p_t)^2\]
            \item Initially, $D_0 = \pm f(1-f)$, with $D$ being positive if $p_0=f$
                and negative if $p_0=1-f$
            \item Together, this forms a nonlinear system of equations for the
                deterministic change in allele frequencies at the two loci (one
                selected, one neutral) and LD between them -- note that this is
                deterministic in the infinite population size limit
        \end{itemize}
    \item Using this model, we can predict how allele frequencies and LD are expected
        to change over time after admixture, for a given effect size $a$ and 
        recombination rate $r$ (Figure~\ref{fig:linkage-pred}A,B)
    \item As expected, lower effect sizes result in a slower reduction in introgressed
        ancestry frequency, while higher recombination rates more quickly decouple
        the linked ancestry from the selected allele
    \item Thus, the expected reduction in introgressed ancestry is largest for
        larger effect sizes (Figure~\ref{fig:linkage-pred}C), and LD between the
        selected and linked neutral alleles is largest for neutral sites closest
        to the selected allele (Figure~\ref{fig:linkage-pred}D)
\end{itemize}

@@

\begin{itemize}
    \item To explore the accuracy of this two-locus model, we compared to
        simulations using the demographic model shown in
        Figure~\ref{fig:toy-admixture}A with introgression fraction $f=0.05$.
        We simulated a single 1 Morgan
        chromosome, with all mutations having effect sizes $\pm a$ ($a=0.02$ or
        $0.05$, with equal probability of being trait-increasing or decreasing)
        and varying the total mutation rate ($\mu=0.001$, $0.0025$, and $0.01$)
    \item For each mutation that was fixed in one population and absent in the
        other (either fixed in the minor or major ancestry population), we
        determined the average introgressed ancestry surrounding those introgressed
        alleles (Figure~\ref{fig:linkage-sim})
    \item The deterministic model provides a very good approximation for linked
        introgressed ancestry when the mutation rate is small
    \item As the mutation rate increases, polygenicity increases so that
        trait-affecting alleles are more densely distributed along the chromosome --
        deviations from the deterministic model are due to interference between
        tightly linked trait-affecting alleles. For the highest mutation rate shown
        here, there will be many trait-affecting alleles within the 1 cM window
        around any focal SNP
\end{itemize}

@@

\subsection*{Introgressed ancestry deserts are shared under reciprocal introgression}

>>>>

\subsection*{The effects of linkage}

@@

The strength of the deviation of $V_G$ between models with and without linkage
depends on a number of factors. The total mutation rate affects not only the
polygenicity, with higher mutation rates causing higher number of segregating
alleles, but also the mean recombination rate between those alleles, as they
will be more or less densely distributed in the genome. The distribution of
effect sizes plays an important role, as seen by the opposite trends in $V_G$
relative to unlinked predictions (Figure~X vs. Figure~Y), which are most
apparent with high mutation rates, and thus high polygenicity. In such cases,
there appear to be complex dynamics involving the Bulmer Effect
\citep{bulmer1971effect} and interference between selected alleles
\citep{hill1966effect}.

@@

Because introgressed trait-affecting alleles are selection against as the minor
allele, introgressed ancestry segments in the regions surrounding the selected
alleles will also be removed due to linkage. The expected reduction in such
introgressed ancestry will depend on the effect size of the trait-affecting
allele and the probability of recombination surrounding that locus. We first
consider a deterministic model of the dynamics of introgressed allele and
linked ancestry frequencies with variable recombination between them. This
simple model ignores the effects of drift and of interference between multiple
selected alleles.

We model a trait-affecting site with a fixed difference between the two
parental populations, in which the derived allele may be fixed in either
population. At the functional site, with admixture proportion $f$ from the
minor parental source, the initial frequency of the derived allele is either
$p_0=f$ or $p_0=1-f$. Over one generation, the expected allele frequency at the
selected locus is, to leading order in $s$,
\begin{align}\label{eq:system-p}
    p_{t+1} & = p_t - s p_t(1-p_t)(1-2p_t),
\end{align}
with
\(s=a^2/2V_S\). As discussed above, if \(p_0<1/2\), \(p_t\rightarrow0\), and if
\(p_0>1/2\), \(p_t\rightarrow1\) as \(t\rightarrow\infty\).

We consider a neutral locus separated from the selected locus by recombination
probability $r$. Initially, the expected frequency of linked introgressed
ancestry is \(q_0=f\), which changes over time due to linked selection on the
trait-affecting allele. Letting \(D=Cov(p,q)\) be the standard covariance
measure of LD between the two loci, $q$ is expected to change as
\begin{align}\label{eq:system-q}
    q_{t+1} & = q_t - s D_t(1-2p_t).
\end{align}
Initially, \(D_0=\pm f(1-f)\), with \(D\) being
positive if \(p_0=f\) and negative if \(p_0=1-f\). LD between the loci changes
deterministically over time, due to both selection and recombination, so that
\begin{align}\label{eq:system-D}
    D_{t+1} = D_t - r D_t - s D_t (1-2p_t)^2.
\end{align}
Together, this forms a
simple nonlinear system of equations for the deterministic change in allele
frequencies at the two loci and LD between them.

Using this model, we predict the changes in introgressed allele frequences and
LD after admixture for given effect size $a$ and recombination rate $r$
(Figure~\ref{fig:linkage-pred}). As expected, smaller effect sizes result in a
slower decay in introgressed ancestry frequency at both selected and linked
loci, and larger recombination rates more quickly decouple the linked ancestry
from the selected allele dynamics. Thus, the expected reduction in introgressed
ancestry is largest for larger effect sizes (Figure~\ref{fig:linkage-pred}X),
and LD between the selected and linked neutral alleles is largest for neutral
sites closest to the selected allele (Figure~\ref{fig:linkage-pred}Y).

We assessed the accuracy of the deterministic two-locus model using
individual-based forward-in-time simulations \citep{thornton2019polygenic}
under a simple demographic model (Figure~\ref{fig:toy-admixture}A), with
introgression fraction \(f=0.05\). We simulated a single chromosome of length 1
Morgan, with all mutations having effect sizes \(\pm a\) (\(a=0.02\) or
\(0.05\)), and we varied the total per-chromosome mutation rate (\(\mu=0.001\),
\(0.0025\) and \(0.01\)). For each fixed-difference mutation in the parental
populations, we determined the average introgressed ancestry surrounding such
loci (Methods).

The deterministic model (Equations~\ref{eq:system-p}--\ref{eq:system-D})
provides a very good approximation when mutation rates are small. As the
mutation rate increases in these simulations, polygenicity also increases so
that trait-affecting alleles are more densely distributed along the chromosome.
Deviations from the deterministic model are due to multiple selected alleles
affecting local introgressed ancestry. For the highest mutation rate shown
here, there are expected to be multiple other trait-affecting alleles within a
1 cM window around any focal SNP, which interact with each other and also
affect local ancestry proportions.

@@

@@

\subsection*{Introgressed ancestry deserts are shared under stabilizing
selection and reciprocal introgression}

As shown in the previous section, introgressed ancestry is reduced around loci
with trait-affecting alleles regardless of the parental population the derived
allele is present in. We should therefore expect to observe reductions in
introgressed ancestry at the same trait-affecting loci after gene flow in
either direction. If introgression occurs in both directions, regions of
reduced introgressed ancestry will coincide around such loci, and introgression
``deserts'' that appear due to selection against trait-affecting alleles will
tend to be shared.

To demonstrate this effect, we performed chromosome-scale simulations under a
model of human-Neanderthal reciprocal introgression
\citep[Figure~\ref{fig:neand-to-human}A,][]{harris2023diverse}. Mutations
affecting a trait under stabilizing selection in each population could arise in
100 kb ``functional'' regions, spaced 1 Mb apart (Methods). Sampling
individuals from both the human and Neanderthal lineages, we observed the
average introgressed ancestry proportions were lowest within functional regions
and dipped in the surrounding regions (Figure~\ref{fig:deserts}A). This
corresponded to an increased proportion of ancestry deserts (defined as
    contiguous 50kb regions with no observed introgressed ancestry in the
sample) within and immediately surrounding the functional regions. Functional
regions also displayed an enrichment of \emph{shared} ancestry deserts, with
the probability of observing ancestry deserts (either within a population or
shared) decaying to background levels as the distance from the functional
region increases.
